---
title: "EDA-Recruiting Case"
output: html_notebook
---

```{r}
# Loading necessary packages and datasets
library(readr)
library(readxl)
library(stringr)
library(lubridate)
library(ggplot2)
library(data.table)
library(plyr)
library(dplyr)
library(scales)

data = read_excel("Case #2 - USC Recruiting Strategy Case - DATA - sept 2018.xlsx")
codebook = read_excel("CASE #2 - Recruiting Strategy CodeBook9.5.18.xlsx")

# Replacing all special character in the column names to underline "_" or simply remove them
colnames(data) = gsub("-", "_", colnames(data))
colnames(data) = gsub(" ", "_", colnames(data))
colnames(data) = gsub("/", "_", colnames(data))
colnames(data) = gsub("\\(", "", colnames(data))
colnames(data) = gsub("\\)", "", colnames(data))

codebook$Field = gsub("-", "_", codebook$Field)
codebook$Field = gsub(" ", "_", codebook$Field)
codebook$Field = gsub("\\(", "", codebook$Field)
codebook$Field = gsub("\\)", "", codebook$Field)

# Replace "-" in "sip_score" and "sip_comments" with NA
data$sip_comments[data$sip_comments == "-"] = NA
data$sip_score[data$sip_score == "-"] = NA
```

## PART 1: Dataset Overview
```{r}
# Creating summary statistics table
summary = data.frame(Variable = colnames(data))
summary$DataType = sapply(data, class)                             # Creating column "DataType" to indicate all variables' data type
summary$NAnum = sapply(data, function(x){sum(is.na(x))})           # Creating column "NAnum" to count the number of NAs in each variable
summary$uniqueNum = sapply(data, function(x){length(unique(x))})   # Creating column "uniqueNum" to count the number of unique values in each variable
summary$min = sapply(data, min, na.rm = TRUE)                      # Creating column "min" to indicate all variables' minimum value (only meaningful for numeric values)
summary$max = sapply(data, max, na.rm = TRUE)                      # Creating column "max" to indicate all variables' maximum value (only meaningful for numeric values)
summary$mean = sapply(data, mean, na.rm = TRUE)                    # Creating column "meab" to indicate all variables' mean (only meaningful for numeric values)

# Join variable definitions in codebook to summary statistics table
summary = merge(summary, codebook, by.x = "Variable", by.y = "Field")

# List 5 examples of the field values for each variable
example = sapply(data, function(x){sort(unique(x)[1:5])})         # extract and sort unique values for each column
example = ldply(example, rbind)                                   # transform the list into a dataframe
colnames(example) = c("Variable", "e1", "e2", "e3", "e4", "e5")   # rename the columns
example$examples = apply(example[,-1], 1, paste, collapse = "//") # paste all the 5 examples for each variable into one column
summary = merge(summary, example[, c(1, 7)], by = "Variable")     # join the examples to the summary table

# Write the summary statistics into csv file
#fwrite(summary, "summary.csv")
```

Since our business objective only deals with 1st tier admitted students, we will only perform cleaning and exploratory analysis on the 1st tier admitted students.
```{r}
t1 = data%>%
  filter(decision == "1st Tier Admit") #not used. Reasons stated below
```


## PART 2: Exploring Data Distribution
There are altogether 81 variables in the dataset. We will go over each column one by one to understand its details and then decide the corresponding cleaning method.

1. opportunity_id: the unique identifer of each applicant. 
2. application_term: Term applied to program
There is a sudden drop in number of applications for 1st tier admit in 2015, so I checked the whole dataset to see if the sudden drop is due to the drop in total number of application, and find that the drop is universal among all students. --> OUR BUSINESS OBJECTIVE MAY NOT BE FEASIBLE!
```{r}
# class(t1$application_term) #data type is character
# 
# # convert the data type to factor and reorder the level to match actual semesters timeline
# t1$application_term = as.factor(t1$application_term)
# levels(t1$application_term) = levels(t1$application_term)[c(7, 1, 8, 2, 9, 3, 10, 4, 5, 6)] 
# levels(t1$application_term)
# 
# # draw the plot
# ggplot(t1, aes(x = application_term))+
#   geom_bar(fill="#999999")+
#   labs(title="Distribution of application_term", 
#        x="application_term", 
#        y="Count")+
#   theme(plot.title = element_text(hjust = 0.5),
#         axis.text.x = element_text(angle = 45, hjust = 1))

# convert the data type to factor and reorder the level to match actual semesters timeline
data$application_term = as.factor(data$application_term)
levels(data$application_term) = levels(data$application_term)[c(10, 1, 11, 2, 7, 12, 3, 8, 13, 4, 5, 9, 14, 6)]

# draw the plot (grey is all students, red is 1st tier students)
data%>%
  filter(!is.na(application_term))%>%
  ggplot(aes(x = application_term))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of application_term (for all applications)", 
       x="application_term", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))+
  geom_bar(data = t1, fill="red")+
  labs(title="Distribution of application_term (for all applications)", 
       x="application_term", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))

# draw the count for students who confirmed offer
data%>%
  filter(response == "Confirmed")%>%
  ggplot(aes(x = application_term))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of application_term (for confirmed stduents)", 
       x="application_term", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
So we performed the rest of the EDA, on the entire dataset
3. degree_posted: When their undergraduate degree conferred
1 outlier 
```{r}
# draw the plot
data%>%
  filter(degree_posted < "2019-01-01")%>%
  ggplot(aes(x = degree_posted))+
  geom_point(stat = "count")+
  labs(title = 'Distribution of undergraduate degree conferred date')+
  theme(plot.title = element_text(hjust = 0.5),
         text = element_text(size=15))
```

4. application_received_date: Date their application was received
many NAs
```{r}
# aggregated
data%>%
  filter(!is.na(application_received_date))%>%
  ggplot(aes(x = as.factor(month(application_received_date))))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of application_received_date (agg. to months)',
       x = "month")+
  theme(plot.title = element_text(hjust = 0.5),
         text = element_text(size = 15))

# across all years
data%>%
  filter(!is.na(application_received_date))%>%
  ggplot(aes(x = application_received_date))+
  geom_point(stat = "count")+
  labs(title = 'Distribution of application_received_date',
       x = "time")+
  theme(plot.title = element_text(hjust = 0.5),
        text = element_text(size = 15))
```

5. file_status_completed_date: Date their application file was completed

```{r}
# aggregated to months
data%>%
  filter(file_status_completed_date < "2019-01-01")%>%
  ggplot(aes(x = as.factor(month(file_status_completed_date))))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of file_status_completed_date (agg. to months)',
       x = "month")+
  theme(plot.title = element_text(hjust = 0.5),
         text = element_text(size = 15))

# across all years
data%>%
  filter(file_status_completed_date < "2019-01-01")%>%
  ggplot(aes(x = file_status_completed_date))+
  geom_point(stat = "count")+
  labs(title = 'Distribution of file_status_completed_date',
       x = "time")+
  theme(plot.title = element_text(hjust = 0.5),
         text = element_text(size = 15))
```

6. decision_date: Date of the admit or deny decision for their application
There is one outlier in this field and is apparently a typo, which is excluded from the dataset
```{r}
# draw the plot
data%>%
  filter(decision_date < "2019-01-01")%>%
  ggplot(aes(x = as.factor(month(decision_date))))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of decision_date',
       x = "month")+
  theme(plot.title = element_text(hjust = 0.5),
         text = element_text(size = 15))

# across all years
data%>%
  filter(decision_date < "2019-01-01")%>%
  ggplot(aes(x = decision_date))+
  geom_point(stat = "count")+
  labs(title = 'Distribution of decision_date',
       x = "time")+
  theme(plot.title = element_text(hjust = 0.5),
         text = element_text(size = 15))
```

7. application_cleaned_date: Date their application was processed after receiving
```{r}
# draw the plot
data%>%
  filter(!is.na(application_cleaned_date))%>%
  ggplot(aes(x = as.factor(month(application_cleaned_date))))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of application_cleaned_date',
       x = "month")+
  theme(plot.title = element_text(hjust = 0.5),
         text = element_text(size = 15))

# across all years
data%>%
  filter(!is.na(application_cleaned_date))%>%
  ggplot(aes(x = application_cleaned_date))+
  geom_point(stat = "count")+
  labs(title = 'Distribution of application_cleaned_date',
       x = "time")+
  theme(plot.title = element_text(hjust = 0.5),
         text = element_text(size = 15))
```

8. deposit_received_date
```{r}
# draw the plot
data%>%
  filter(!is.na(deposit_received_date))%>%
  ggplot(aes(x = as.factor(month(deposit_received_date))))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of deposit_received_date',
       x = "month")+
  theme(plot.title = element_text(hjust = 0.5),
         text = element_text(size = 15))

# across all years
data%>%
  filter(!is.na(deposit_received_date))%>%
  ggplot(aes(x = deposit_received_date))+
  geom_point(stat = "count")+
  labs(title = 'Distribution of deposit_received_date',
       x = "time")+
  theme(plot.title = element_text(hjust = 0.5),
         text = element_text(size = 15))
```

9. department: Department of students
```{r}
# draw the plot and exclude all NAs (many NAs)
data%>%
  filter(!is.na(department))%>%
  ggplot(aes(x = department))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of department')+
  theme(plot.title = element_text(hjust = 0.5),
         text = element_text(size = 15))
```

10. undergraduate_gpa: calculated by USC
```{r}
#plot
data%>%
  filter(undergraduate_gpa != 0)%>%
  ggplot(aes(x = undergraduate_gpa))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of undergraduate_gpa')+
  theme(plot.title = element_text(hjust = 0.5),
         text = element_text(size = 15))
```

11. self_reported_gpa
Outliers with self reported gpa greater than 5 and equals to 0 (should be cleaned)
```{r}
#plot
data%>%
  filter(self_reported_gpa <= 4 & self_reported_gpa != 0)%>%
  ggplot(aes(x = self_reported_gpa))+
  geom_point(stat = "count")+
  labs(title = 'Distribution of self_reported_gpa')+
  theme(plot.title = element_text(hjust = 0.5),
         text = element_text(size = 15))

#plot self_reported_gpa with undergraduate gpa calculted by USC
data%>%
  filter(self_reported_gpa <= 4 & self_reported_gpa != 0)%>%
  filter(undergraduate_gpa != 0)%>%
  ggplot(aes(x = undergraduate_gpa))+
  geom_bar(fill = "#999999")+
  geom_point(aes(x = self_reported_gpa), stat = "count", color = "red")+
    labs(title = 'self_reported_gpa (red) against undergraduate_gpa')+
  theme(plot.title = element_text(hjust = 0.5),
         text = element_text(size = 15))
```

12. first_reviewer_id
```{r}
#plot
data%>%
  filter(!is.na(first_reviewer_id))%>%
  ggplot(aes(x = first_reviewer_id))+
  geom_point(stat = "count")+
  labs(title = 'Distribution of first_reviewer_id')+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

13. first_review_exp: First reviewer score for experience
```{r}
data%>%
  ggplot(aes(x = first_review_exp))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of first_review_exp')+
  theme(plot.title = element_text(hjust = 0.5))
```

14. first_review_sop: First reviewer score for statement of purpose
```{r}
data%>%
  ggplot(aes(x = first_review_sop))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of first_review_sop')+
  theme(plot.title = element_text(hjust = 0.5))
```

15. first_review_gpa: First reviewer score for GPA (undergrad and grad)
```{r}
data%>%
  ggplot(aes(x = first_review_gpa))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of first_review_gpa')+
  theme(plot.title = element_text(hjust = 0.5))
```

16. first_review_lor: letters of recommendation
```{r}
data%>%
  ggplot(aes(x = first_review_lor))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of first_review_lor')+
  theme(plot.title = element_text(hjust = 0.5))
```

17. first_review_total: total score
```{r}
data%>%
  ggplot(aes(x = first_review_total))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of first_review_total')+
  theme(plot.title = element_text(hjust = 0.5))
```

18. second_reviewer_id
```{r}
#plot
data%>%
  filter(!is.na(second_reviewer_id))%>%
  ggplot(aes(x = second_reviewer_id))+
  geom_point(stat = "count")+
  labs(title = 'Distribution of second_reviewer_id')+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

19. second_review_exp: second reviewer score for experience
```{r}
data%>%
  ggplot(aes(x = second_review_exp))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of second_review_exp')+
  theme(plot.title = element_text(hjust = 0.5))
```

20. second_review_sop: second reviewer score for statement of purpose
```{r}
data%>%
  ggplot(aes(x = second_review_sop))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of second_review_sop')+
  theme(plot.title = element_text(hjust = 0.5))
```

21. second_review_gpa: second reviewer score for GPA (undergrad and grad)
```{r}
data%>%
  ggplot(aes(x = second_review_gpa))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of second_review_gpa')+
  theme(plot.title = element_text(hjust = 0.5))
```

22. concentration/department_interest: Department of student. 
This one is the unofficial one ("department" is the official one), probably manually filled and thus is very messy. We can directly use 

23. first_review_lor: letters of recommendation
```{r}
data%>%
  ggplot(aes(x = second_review_lor))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of second_review_lor')+
  theme(plot.title = element_text(hjust = 0.5))
```

24. second_review_total: total score
```{r}
data%>%
  ggplot(aes(x = second_review_total))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of second_review_total')+
  theme(plot.title = element_text(hjust = 0.5))
```

25. billing_state/province
Manually input data and thus very messy. Lots of inconsistent writing of same geographical location, such as "CA" and "California".

26. billing_zip/postal_code

27. billing_country
```{r}
data$billing_country[data$billing_country %in% c("united States", "United", "USA", "United States of America", "CA")] = "United States"

data%>%
  filter(!is.na(billing_country))%>%
  group_by(billing_country)%>%
  summarise(count = n())%>%
  ggplot(aes(x = reorder(billing_country, -count), y = count))+
  geom_bar(stat = "identity", fill = "#999999")+
  labs(title = 'Distribution of billing_country')+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

28. distance_to_campus
```{r}
data%>%
  ggplot(aes(x = distance_to_campus))+
  geom_point(stat = "count")+
  labs(title = 'Distribution of distance_to_campus')+
  theme(plot.title = element_text(hjust = 0.5))
```

29. program_length
```{r}
data$program_length = as.factor(data$program_length)
levels(data$program_length) = levels(data$program_length)[c(2, 4, 6, 7, 1, 8, 3, 5)]

data%>%
  filter(!is.na(program_length))%>%
  ggplot(aes(x = program_length))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of program_length')+
  theme(plot.title = element_text(hjust = 0.5))
```

30. student_id (can be used as the unique identifier as "opportunity id")

31. post_code: meaning program of study
```{r}
data%>%
  filter(!is.na(post_code) & !post_code %in% c("Fall 2015", "1645;264", "264;1412", "264;1411", "264;264", "272;264"))%>%
  group_by(post_code)%>%
  summarise(count = n())%>%
  ggplot(aes(x = reorder(post_code, -count), y = count))+
  geom_bar(stat = "identity", fill = "#999999")+
  labs(title = 'Distribution of post_code',
       x = 'post_code')+
  theme(plot.title = element_text(hjust = 0.5))
```

32. post_code_expired (whether the applicant switched from another program)
```{r}
data%>%
  ggplot(aes(x = post_code_expired))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of post_code_expired')+
  theme(plot.title = element_text(hjust = 0.5))
```

33. file_status
```{r}
data%>%
  ggplot(aes(x = file_status))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of file_status')+
  theme(plot.title = element_text(hjust = 0.5))

#remove NA
data%>%
  filter(!is.na(file_status))%>%
  ggplot(aes(x = file_status))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of file_status')+
  theme(plot.title = element_text(hjust = 0.5))
```

34. birth_date: dob of the applicant
```{r}
data%>%
  filter(!is.na(birth_date) & birth_date < "2000-01-01")%>%
  ggplot(aes(x = as.factor(year(birth_date))))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of birth_date',
       x = 'birth_date (year)')+
  scale_x_discrete(breaks = seq(1930, 2000, by = 10))+
  theme(plot.title = element_text(hjust = 0.5),
         text = element_text(size=15))
```

35. ethnicity
```{r}
data%>%
  filter(!is.na(ethnicity))%>%
  group_by(ethnicity)%>%
  summarise(count = n())%>%
  ggplot(aes(x = reorder(ethnicity, -count), y = count))+
  geom_bar(stat = "identity", fill = "#999999")+
  labs(title = 'Distribution of ethnicity')+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

36. gender
```{r}
data%>%
  filter(!is.na(gender))%>%
  ggplot(aes(x = gender))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of gender')+
  theme(plot.title = element_text(hjust = 0.5))
```

37. campus: Campus of study of applicant
```{r}
data%>%
  filter(!is.na(campus))%>%
  ggplot(aes(x = campus))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of campus')+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

38. first_generation
```{r}
data%>%
  filter(!is.na(first_generation))%>%
  ggplot(aes(x = first_generation))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of first_generation')+
  theme(plot.title = element_text(hjust = 0.5))
```

39. post_code_account
```{r}
data%>%
  filter(!is.na(post_code) & !post_code %in% c("Fall 2015", "1645;264", "264;1412", "264;1411", "264;264", "272;264"))%>%
  group_by(post_code_account)%>%
  summarise(count = n())%>%
  ggplot(aes(x = reorder(post_code_account, -count), y = count))+
  geom_bar(stat = "identity", fill = "#999999")+
  labs(title = 'Distribution of post_code_account',
       x = 'post_code_account')+
  theme(plot.title = element_text(hjust = 0.5))
```

38. registered_units_account
```{r}
data%>%
  filter(!is.na(registered_units_account))%>%
  ggplot(aes(x = registered_units_account))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of registered_units_account')+
  theme(plot.title = element_text(hjust = 0.5))
```

40. department_student_account
```{r}
data%>%
  filter(!is.na(department_student_account))%>%
  ggplot(aes(x = department_student_account))+
  geom_bar(fill = "#999999")+
  labs(title = 'Distribution of department_student_account')+
  theme(plot.title = element_text(hjust = 0.5))
```

