---
title: "EDA-Recruiting Case"
output: html_notebook
---

```{r}
# Loading necessary packages and datasets
library(readr)
library(readxl)
library(stringr)
library(lubridate)
library(ggplot2)
library(data.table)
library(plyr)
library(dplyr)

data = read_excel("Case #2 - USC Recruiting Strategy Case - DATA - sept 2018.xlsx")
codebook = read_excel("CASE #2 - Recruiting Strategy CodeBook9.5.18.xlsx")
```

## PART 1: Dataset Overview
```{r}
# Replacing all spaces in the column names to underline "_"
colnames(data) = gsub(" ", "_", colnames(data))
codebook$Field = gsub(" ", "_", codebook$Field)

# Replacing "NA" in sip scores and comments to NA
data$sip_comments[data$sip_comments == "NA"] = NA
data$sip_score[data$sip_score == "NA"] = NA

# Creating summary statistics table
summary = data.frame(Variable = colnames(data))
summary$DataType = sapply(data, class)                             # Creating column "DataType" to indicate all variables' data type
summary$NAnum = sapply(data, function(x){sum(is.na(x))})           # Creating column "NAnum" to count the number of NAs in each variable
summary$uniqueNum = sapply(data, function(x){length(unique(x))})   # Creating column "uniqueNum" to count the number of unique values in each variable
summary$min = sapply(data, min, na.rm = TRUE)                      # Creating column "min" to indicate all variables' minimum value (only meaningful for numeric values)
summary$max = sapply(data, max, na.rm = TRUE)                      # Creating column "max" to indicate all variables' maximum value (only meaningful for numeric values)
summary$mean = sapply(data, mean, na.rm = TRUE)                    # Creating column "meab" to indicate all variables' mean (only meaningful for numeric values)

# Join variable definitions in codebook to summary statistics table
summary = merge(summary, codebook, by.x = "Variable", by.y = "Field")

# List 5 examples of the field values for each variable
example = sapply(data, function(x){sort(unique(x)[1:5])})         # extract and sort unique values for each column
example = ldply(example, rbind)                                   # transform the list into a dataframe
colnames(example) = c("Variable", "e1", "e2", "e3", "e4", "e5")   # rename the columns
example$examples = apply(example[,-1], 1, paste, collapse = "//") # paste all the 5 examples for each variable into one column
summary = merge(summary, example[, c(1, 7)], by = "Variable")     # join the examples to the summary table

# Write the summary statistics into csv file
fwrite(summary, "summary.csv")
```

## PART 2: Exploring Data Distribution and Data Cleaning
There are altogether 81 variables in the dataset. We will go over each column one by one to understand its details and then decide the corresponding cleaning method
```{r}
# # Filter out first-tier applicants
# t1 <- data %>%
#   filter(decision == "1st Tier Admit")

```
### Distribution for variables
# 43 response
```{r}
ggplot(data, aes(x=response))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of Response", 
       x="Response", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 44 academic_condition
```{r}
ggplot(data, aes(x=academic_condition))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of Academic_Condition", 
       x="Academic Condition", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 45 response_date
```{r}
ggplot(data, aes(x=response_date)) +
  geom_point(stat = "count")+
  labs(title="Distribution of Response Date", 
       x="Response Date", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 46 date_acceptance_letter_sent
# Could create the length of time a student respond to his/her offer letter using variable 45 and 46
```{r}
ggplot(data, aes(x=date_acceptance_letter_sent)) +
  geom_point(stat = "count")+
  labs(title="Distribution of Date_Acceptance_Letter_Sent", 
       x="Date_Acceptance_Letter_Sent", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 47 campus_-_cleaned
```{r}
ggplot(data, aes(x=`campus_-_cleaned`))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of Campus_Cleaned", 
       x="Campus", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 48 Enrolled_Post_add/drop
# Not useful
```{r}
ggplot(data, aes(x=`enrolled_post_ad/drop`))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of Enrolled_Post_add/drop", 
       x="Enrolled_Post_add/drop", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 49 sip_score (remove all NAs in this variable)
```{r}
data%>%
  filter(!is.na(sip_score))%>%
  ggplot(aes(x=sip_score))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of SIP Score", 
       x="SIP Score", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 51 date_sent_to_first_review
```{r}
ggplot(data, aes(x=date_sent_to_first_review)) +
  geom_point(stat = "count")+
  labs(title="Distribution of Date_Sent_To_First_Review", 
       x="Date_Sent_To_First_Review", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 52 date_returned_from_first_review
# Could create the length of first review time using variable 51 and 52
```{r}
ggplot(data, aes(x=date_returned_from_first_review)) +
  geom_point(stat = "count")+
  labs(title="Distribution of Date_Returned_From_First_Review", 
       x="Date_Sent_To_First_Review", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 53 date_sent_to_second_review
```{r}
ggplot(data, aes(x=date_sent_to_second_review)) +
  geom_point(stat = "count")+
  labs(title="Distribution of Date_Sent_To_Second_Review", 
       x="Date_Sent_To_Second_Review", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 54 date_returned_from_second_review (remove one obvious outlier)
# Could create the length of first review time using variable 53 and 54
```{r}
data %>%
  filter(date_returned_from_second_review < "2020-01-01 UTC")%>%
  ggplot(aes(x=date_returned_from_second_review)) +
  geom_point(stat = "count")+
  labs(title="Distribution of Date_Returned_From_Second_Review", 
       x="Date_Sent_To_Second_Review", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 55 residency
```{r}
data$residency = as.factor(data$residency)
levels(data$residency) = c("US Citizen","International","Non Resident","Non Resident","None",
                           "Other","Permanent US Resident", "Permanent US Resident",
                           "Temporary US Resident","US Citizen","US Citizen")
ggplot(data, aes(x=residency))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of Residency", 
       x="Residency", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 56 admission_advisor_assignment
```{r}
ggplot(data, aes(x=admission_advisor_assignment))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of Admission_Advisor_Assignment", 
       x="Admission_Advisor_Assignment", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 57 admissions_term_code (01 = Spring, 02 = Summer, 03 = Fall)
```{r}
data$admissions_term_code = as.numeric(data$admissions_term_code)
data%>%
  filter(admissions_term_code >= 201302)%>%
  mutate(admissions_term_code = as.factor(admissions_term_code))%>%
  ggplot(aes(x=admissions_term_code))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of Admissions_Term_Code", 
       x="Admissions_Term_Code", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 58 first_enrolled_code
```{r}
data$first_enrolled_code = as.numeric(data$first_enrolled_code)
data%>%
  filter(first_enrolled_code >= 201302)%>%
  mutate(first_enrolled_code = as.factor(first_enrolled_code))%>%
  ggplot(aes(x=first_enrolled_code))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of First_Enrolled_Code", 
       x="First_Enrolled_Code", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# Comparing admissions_term_code and first_enrolled_code
```{r}
data%>%
  filter(first_enrolled_code >= 201302 & admissions_term_code >= 201302)%>%
  mutate(first_enrolled_code = as.factor(first_enrolled_code),
         admissions_term_code = as.factor(admissions_term_code))%>%
  select(first_enrolled_code,admissions_term_code)%>%
  melt(measure.vars = c("first_enrolled_code", "admissions_term_code"))%>%
  ggplot (aes(x=value, fill=variable)) + 
  geom_bar (position = position_dodge())+
  labs(title="Distribution of First_Enrolled_Code and Admissions_Term_Code", 
       x="Value", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 59 last_enrolled_code
```{r}
data$last_enrolled_code = as.numeric(data$last_enrolled_code)
data%>%
  filter(last_enrolled_code >= 201302)%>%
  mutate(last_enrolled_code = as.factor(last_enrolled_code))%>%
  ggplot(aes(x=last_enrolled_code))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of Last_Enrolled_Code", 
       x="Last_Enrolled_Code", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 60 first_enrolled_term (Almost identical to first_enrolled_code, has 12 more data points)
```{r}
data$first_enrolled_term = as.numeric(data$first_enrolled_term)
data%>%
  filter(first_enrolled_term >= 201302)%>%
  mutate(first_enrolled_term = as.factor(first_enrolled_term))%>%
  ggplot(aes(x=first_enrolled_term))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of First_Enrolled_Term", 
       x="First_Enrolled_Term", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 61 last_enrolled_term(Almost identical to last_enrolled_code, has 11 more data points)
```{r}
data$last_enrolled_term = as.numeric(data$last_enrolled_term)
data%>%
  filter(last_enrolled_term >= 201302)%>%
  mutate(last_enrolled_term = as.factor(last_enrolled_term))%>%
  ggplot(aes(x=last_enrolled_term))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of Last_Enrolled_Term", 
       x="Last_Enrolled_Term", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 62 current_term
```{r}
ggplot(data,aes(x=current_term))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of Current_Term", 
       x="Current_Term", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 63 processing_complete_date
```{r}
ggplot(data, aes(x=processing_complete_date)) +
  geom_point(stat = "count")+
  labs(title="Distribution of Processing_Complete_Date", 
       x="Processing_Complete_Date", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 64 date_sent_for_gia_review
# # Could create the length of time between data sent for gia review and processing complete date using variable 63 and 64
```{r}
ggplot(data, aes(x=date_sent_for_gia_review)) +
  geom_point(stat = "count")+
  labs(title="Distribution of Date_Sent_for_GIA_Review", 
       x="Date_Sent_for_GIA_Review", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 65 gia_review_date (removing one obvious outlier)
```{r}
data%>%
  filter(gia_review_date < "2020-01-01 UTC")%>%
  ggplot(aes(x=gia_review_date)) +
  geom_point(stat = "count")+
  labs(title="Distribution of GIA_Review_Date", 
       x="GIA_Review_Date", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 66 age
```{r}
data %>%
  filter(age>=15)%>%
  ggplot(aes(x=age))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of Age", 
       x="Age", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 67 undergrad_institution_(self_reported)
```{r}
data%>%
  group_by(`undergrad_institution_(self_reported)`)%>%
  summarise(Count=n())%>%
  arrange(-Count)
```
# 68,69, 70 Text variables, could not be used

# 71 deposit_received
```{r}
ggplot(data,aes(x=deposit_received))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of Deposit_Received", 
       x="Deposit_Received", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 72 ready_for_director_review
```{r}
ggplot(data,aes(x=ready_for_director_review))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of Ready_For_Director_Review", 
       x="Ready_For_Director_Review", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 73 follow_up_required
```{r}
ggplot(data,aes(x=follow_up_required))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of Follow_Up_Required", 
       x="Follow_Up_Required", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 74 90041_mi
```{r}
ggplot(data,aes(x=`90041_mi`))+
  geom_histogram(fill="#999999")+
  labs(title="Distribution of 90041_Mi", 
       x="90041_Mi", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 75 91601_mi
```{r}
ggplot(data,aes(x=`91601_mi`))+
  geom_histogram(fill="#999999")+
  labs(title="Distribution of 91601_Mi", 
       x="91601_Mi", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 76 92122_mi
```{r}
ggplot(data,aes(x=`92122_mi`))+
  geom_histogram(fill="#999999")+
  labs(title="Distribution of 92122_Mi", 
       x="92122_Mi", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 77 94607_mi
```{r}
ggplot(data,aes(x=`94607_mi`))+
  geom_histogram(fill="#999999")+
  labs(title="Distribution of 94607_Mi", 
       x="94607_Mi", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 79 last_60/90_gpa
```{r}
ggplot(data,aes(x=`last_60/90_gpa`))+
  geom_histogram(fill="#999999")+
  labs(title="Distribution of Last_60/90_GPA", 
       x="Last_60/90_GPA", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
# 80 graduate_gpa
```{r}
ggplot(data,aes(x=graduate_gpa))+
  geom_histogram(fill="#999999")+
  labs(title="Distribution of Graduate_GPA", 
       x="Graduate_GPA", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
 # 81 temperature
```{r}
data[!is.na(data$temperature),]%>%
  ggplot(aes(x=temperature))+
  geom_bar(fill="#999999")+
  labs(title="Distribution of Temperature", 
       x="Temperature", 
       y="Count")+
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

